---
title: "ML2 Project Report"
author: "Nick Chandler, Tharun Johny Mekala, Ashritha Shreedhar, Luke Richard"
date: "2025-12-31"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
  html_document: default
bibliography: references.bib
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(dplyr)
library(corrplot)
library(glmnet)
library(iml)
library(dplyr)
library(lightgbm)
library(ggplot2)
library(gridExtra)
library(patchwork)
```

# Abstract
In this report, we examine the application of two statistical learning methods to baseball data from the years 2020-2024. Specifically we are interested in predicting the next season's batting average of arbitrary players given the standard and advanced metrics of the previous season from this [source](https://www.fangraphs.com/leaders/major-league). The report begins with a quick exploratory data analysis (EDA), it then moves into a mathematical description of the statistical learning methods applied, then the application of each method, and finally a performance evaluation and description. This report illustrates how classical, interpretable methods can be applied, and then how more modern (and less interpretable) methods can be applied for better performance, and with the help of XAI, also be interpretable.

# EDA
Here we do a short exploratory analysis of the data to understand some of the basic characteristics with a focus primarily on the correlation structures between variables since the modeling techniques we utilize rely on this. For more information on each feature, fangraphs provides a glossary, [here](https://library.fangraphs.com/offense/offensive-statistics-list/).

```{r, echo=FALSE}
data <- read.csv("merged.csv")
```

```{r, echo=FALSE}
num_unique_players <- data %>%
  distinct(Name) %>%       # keep only unique player names
  nrow()                   # count them
```
The dataset has `r nrow(data)` rows, `r ncol(data)` columns, and contains `r num_unique_players` unique players.

Here are correlation plots using Pearson's correlation coefficient [@pearson1895]. Pay special attention to the *next_AVG* column so see which features have a high (or low) correlation with the target.
```{r, echo = FALSE, fig.width=4, fig.height=4, fig.align='left'}
numeric_data <- data[, sapply(data, is.numeric)]

cor_matrix_pearson <- cor(numeric_data, use = "pairwise.complete.obs", method = "pearson")
corrplot(cor_matrix_pearson, method = "color", tl.cex = 0.6)
```

# Mathematical Description of Methods

This section briefly describes the mathematical background of each of the methods used. All equations herein can be traced back to the literature of [@hastie2009; @course_notes].

## Mathematical Description of Regularized Regression
This section examines the mathematical background for the regularized regression model types of: Ridge, Lasso, and Elastic-Net regression. All assume a model of the form (here, without interaction terms):
$$y = \beta_0 + \sum_{i=1}^{p} \beta_i x_i + \varepsilon.$$
Where the $\beta_i$s are found by solving (though specific instances introduce additional terms):
$$\min_{\vec{\beta}} \left\|  X\vec{\beta} - \vec{y} \right\|^2_2\text{, with } X \in \mathbb{R}^{n \times p} \text{ the design matrix}$$
The general solution is given (analytically) by:
$$\hat{\vec{\beta}} = \left(X^T X \right)^{-1} X^T\vec{y}$$

### Elastic-Net Regression
The elastic-net regression uses a convex combination of the lasso penalty (L1) and ridge penalty (L2). Specifically, the optimization problem is given by:
$$\min_{\vec{\beta}} \left\| X\vec{\beta} - \vec{y} \right\|_2^2 + \lambda \left( \left(1 - \alpha  \right) \| \vec{\beta} \|_1 + \alpha\|\vec{\beta}  \|_2^2 \right)$$
Since the L1 norm is not differentiable at $0$, there is no closed-form, analytical solution to the above problem. Solvers for convex problems capable of dealing with non-smooth objectives must be applied. The optimality condition is given by: 

$$
0 \in 2 X^\top (X \hat{\beta} - y) +  \lambda \left( 2 \alpha \hat{\vec{\beta}} + (1 - \alpha) \partial \|\hat{\beta}\|_1 \right),
\quad
(\partial \|\hat{\beta}\|_1)_i =
\begin{cases}
\{1\}, & \hat{\beta}_i > 0,\\[1mm]
\{-1\}, & \hat{\beta}_i < 0,\\[1mm]
[-1,1], & \hat{\beta}_i = 0
\end{cases}
$$

The parameters $\lambda$ and $\alpha$ are hyperparameters to be tuned via cross validation.


## Mathematical Description of Gradient Boosting Decesion Trees
In this section we give a short mathematical description of Gradient Boosting Decision Trees (focusing on the gradient boosting aspect) [@course_notes].

Suppose our training data is given by: $\left\{ (\vec{x_i}, y_i) \right\}_{i=1}^n$

Gradient boosting builds an additive sequential model interatively, aiming to minimize a loss function: $L(y, F(x))$. Specifically, gradient boosting builds a model (defined recursively):
$$F_m(x) = F_{m-1}(x) + \eta \cdot \gamma_m \cdot h_m(x)$$
with,
$$F_0(x) = \underset{\gamma}{\arg \min} \sum_{i=1}^{n} L (y_i, \gamma)$$
The two step process (for each iteration) to fit the model consists of:
1. Building weak learners $h_m(x)$.
2. Finding the optimal step size at that iteration $\gamma_m$.

For step 1. we must compute the pseudo residual: 
$$\tilde{y}_{im} = - \left[ \frac{ \partial L(y_i, F(x_i)) }{ \partial F(x_i) } \right]_{F(x) = F_{m-1}(x)}$$
and then fit the weak learner $h_m(x)$ (in our case, a tree model) to the given pseudo residual. 

We then do step 2. to find the optimal step size $\gamma_m$ using an optimizer to solve:
$$\underset{\gamma}{\arg \min} \sum_{i = 1}^{n} L(y_i, F_{m-1}(x_i) + \gamma h_m(x_i))$$
Finally, the $\eta$ given in the original equation is a hyperparameter and must be tuned.


# Train-Val-Test Split

We split the data into a train set (60%), validation (val) set (20%), and a test set (20%). Specifically, we adhere to the following condition when deciding on the splits: A player may occur in no more than one set. Additionally, the newest data (that from the 2024 season) is saved for the test set.
```{r, echo=FALSE}
# All unique players
players <- unique(data$Name)

## PLAN: Take all 2024-appearing players and then more to get the 20% necessary for test. Then just split the val and train normally but with players not occurring in two sets.

# Players in 2024 season -> all go to test
players.2024 <- unique(data[data$Season == 2024,"Name"])
players.remaining <- setdiff(players, players.2024)  # setdiff -> set difference

# How many players to get 20% of rows
rows.2024 <- data[data$Name %in% players.2024, ]
target.test.rows <- floor(0.2 * nrow(data))
extra.needed <- target.test.rows - nrow(rows.2024)

# sample extra players from remaining players
extra.players <- c()
cum.rows <- 0
set.seed(123)
shuffled <- sample(players.remaining)  # Shuffles the players.remaining list

for (p in shuffled) {
  player.rows <- nrow(data[data$Name == p, ])  # Take out the player's rows
  cum.rows <- cum.rows + player.rows  # Add to the count of the cumulative rows
  extra.players <- c(extra.players, p)  # Add new player to the extra players
  if (cum.rows >= extra.needed) break  # Stop if we got the needed players (won't overshoot 20% by much)
}

test.players <- c(players.2024, extra.players)
test.set <- na.omit(data[data$Name %in% test.players, ])

trainval.players <- setdiff(players, test.players)
trainval.data <- data[data$Name %in% trainval.players, ]

# val - 20%
val.frac <- 0.247  # MAGIC NUMBER (kind of, at least there's a var name) !
val.players <- sample(unique(trainval.data$Name),
                      size = ceiling(val.frac * length(unique(trainval.data$Name))))
val.set <- na.omit(trainval.data[trainval.data$Name %in% val.players, ])

# Train is the rest
train.set <- na.omit(trainval.data[!trainval.data$Name %in% val.players, ])


```
In numbers:

- The train set has `r nrow(train.set)` rows and `r length(unique(train.set$Name))` unique players.
- The validation set has `r nrow(val.set)` rows and `r length(unique(val.set$Name))` unique players.
- The test set has `r nrow(test.set)` rows and `r length(unique(test.set$Name))` unique players.  


# Regularized Regression Application

In this section, we apply the regularized regression techniques doing cross validated grid-search hyperparameter optimization over values for alpha.

```{r, echo=FALSE}
drop_cols <- c("Season", "Name", "Team", "next_AVG")

x_train <- as.matrix(train.set[, !(names(train.set) %in% drop_cols)])
y_train <- train.set$next_AVG

x_val <- as.matrix(val.set[, !(names(val.set) %in% drop_cols)])
y_val <- val.set$next_AVG

# Define the alpha sets
ridge_alpha <- 0
lasso_alpha <- 1
elastic_alphas <- seq(0.1, 0.9, by = 0.1)  # elastic net grid (exclude 0 and 1)

# Store results
results <- list()

# Function to train and evaluate
train_eval <- function(alpha_value) {
  cv_fit <- cv.glmnet(x_train, y_train, alpha = alpha_value, nfolds = 5)  # Be able to explain cross validation
  pred_val <- predict(cv_fit, s = "lambda.min", newx = x_val)  # You must understand this well - What is lambda.min, what exactly does it do?
  mse_val <- mean((y_val - pred_val)^2)
  list(model = cv_fit, mse = mse_val, alpha = alpha_value)
}

# 1) Ridge
results[["ridge"]] <- train_eval(ridge_alpha)

# 2) Lasso
results[["lasso"]] <- train_eval(lasso_alpha)

# 3) Elastic Net Grid
for (a in elastic_alphas) {
  key <- paste0("elastic_alpha_", a)
  results[[key]] <- train_eval(a)
}

# Find best model overall
best_key_lr <- names(results)[which.min(sapply(results, function(x) x$mse))]
best_model_lr <- results[[best_key_lr]]$model
best_mse_lr <- round(results[[best_key_lr]]$mse, 3)
```

Through our analysis, we find that `r best_key_lr` attains the minimum validation RMSE of all models tested at (rounded): `r round(sqrt(best_mse_lr),6)`. Which means that the model mispredicts the next season's batting average by approximately `r round(sqrt(best_mse_lr)*100,4)`% on average. We examine the best model further, using classical diagnostics.

The features with larger magnitude coefficients can be seen as contributing more to the regression and so we plot them here.
```{r largest-lr-coefs, echo=FALSE}
# Extract coefficients at lambda.min
coef_best <- as.numeric(coef(best_model_lr, s = "lambda.min"))
names_coef <- rownames(coef(best_model_lr))
coef_df <- data.frame(
  Feature = names_coef,
  Coefficient = coef_best
)

# Remove the intercept
coef_df <- coef_df %>% filter(Feature != "(Intercept)")

# Select top 3 by absolute value
top3_coef <- coef_df %>%
  slice_max(order_by = abs(Coefficient), n = 3)

# Display in a kable
kable(top3_coef, digits = 4)
```

The following is a plot of the lambda value in the `r best_key_lr` regression.
```{r, echo=FALSE, fig.width=5, fig.height=4, fig.align='left'}
lambda_star <- best_model_lr$lambda.min
plot(best_model_lr)
abline(v = log(lambda_star), col = "red", lty = 2, lwd = 2)
```

# Gradient Boosted Decision Trees Application

In this section, we apply LightGBM [@lightgbm].

## LightGBM
Here we do grid-search hyperparameter optimization to find the best LightGBM model.
```{r, echo=FALSE}
# lightgbm training and evaluation function
train_eval_lgb <- function(x_train, y_train, x_val, y_val, params, nrounds=5000, early_stop=100) {
  
  dtrain <- lgb.Dataset(data = x_train, label = y_train)
  dval   <- lgb.Dataset(data = x_val, label = y_val)
  
  evals <- list(train = dtrain, val = dval)
  
  model <- lgb.train(
    params = params,
    data = dtrain,
    nrounds = nrounds,
    valids = evals,
    early_stopping_rounds = early_stop,
    verbose = -1
  )
  
  # Predictions on validation set
  preds <- predict(model, x_val, num_iteration = model$best_iter)
  rmse_val <- sqrt(mean((preds - y_val)^2))
  
  # Record train/val RMSE for plotting
  train_rmse <- sapply(model$record_evals$train$rmse$eval, identity)
  val_rmse   <- sapply(model$record_evals$val$rmse$eval, identity)
  
  return(list(
    model = model,
    rmse_val = rmse_val,
    best_iter = model$best_iter,
    train_rmse = train_rmse,
    val_rmse = val_rmse
  ))
}

# function for plotting train/val curves
plot_rmse_curves <- function(train_rmse, val_rmse) {
  df_plot <- data.frame(
    Iteration = 1:length(train_rmse),
    Train_RMSE = train_rmse,
    Val_RMSE = val_rmse
  )
  
  ggplot(df_plot, aes(x = Iteration)) +
    geom_line(aes(y = Train_RMSE, color = "Train")) +
    geom_line(aes(y = Val_RMSE, color = "Validation")) +
    labs(y = "RMSE", title = "Train vs Validation RMSE") +
    scale_color_manual(values = c("Train"="blue", "Validation"="red")) +
    theme_minimal()
  
}

#function for hyperparamter optimization
hpo_lgb <- function(x_train, y_train, x_val, y_val, param_grid, nrounds=5000, early_stop=100) {
  
  best_rmse <- Inf
  best_params <- NULL
  best_model <- NULL
  best_curves <- NULL
  
  for (i in 1:nrow(param_grid)) {
    
    params <- as.list(param_grid[i, ])
    params$objective <- "regression"
    params$metric <- "rmse"
    
    res <- train_eval_lgb(x_train, y_train, x_val, y_val, params, nrounds, early_stop)
    
    
    if (res$rmse_val < best_rmse) {
      best_rmse <- res$rmse_val
      best_params <- params
      best_model <- res$model
      best_curves <- list(train = res$train_rmse, val = res$val_rmse)
    }
  }
  
  return(list(
    best_model = best_model,
    best_params = best_params,
    best_rmse = best_rmse,
    train_rmse = best_curves$train,
    val_rmse = best_curves$val
  ))
}

print_best_lgb <- function(hpo_result) {
  
  cat("=====================================\n")
  cat("          Best LightGBM Model        \n")
  cat("=====================================\n")
  
  cat(sprintf("Best Validation RMSE : %.4f\n", hpo_result$best_rmse))
  cat(sprintf("Best Training RMSE   : %.4f\n", tail(hpo_result$train_rmse, 1)))
  cat(sprintf("Best Iteration       : %d\n", hpo_result$best_model$best_iter))
  cat("\n")
  cat("Selected Hyperparameters:\n")
  
  # Print each parameter neatly
  for (param_name in names(hpo_result$best_params)) {
    cat(sprintf("  %-20s : %s\n", param_name, hpo_result$best_params[[param_name]]))
  }
  
  cat("=====================================\n")
}
```

```{r, echo=FALSE}
# parameter search grid for HPO
param_grid <- expand.grid(
  learning_rate = c(0.001, 0.005, 0.01),
  max_depth = c(5,6,7),
  num_leaves = c(8,16),
  feature_fraction = c(0.8, 1)
)


# run the hpo
best_lgb <- hpo_lgb(x_train, y_train, x_val, y_val, param_grid, nrounds = 1000, early_stop = 50)
```

```{r plot-lgbm-train-val, echo=FALSE, fig.width=4, fig.height=3, fig.align='left'}
# plot train/val curves
plot_rmse_curves(best_lgb$train_rmse, best_lgb$val_rmse)
```

```{r print-optimal-lgbm-hyperparameters, echo=FALSE}
# print best results
print_best_lgb(best_lgb)
```



## LightGBM Feature Importance

This section examines the LightBGM-Specific feature importances.

```{r, echo=FALSE}
# feature importance
lgb_importance <- lgb.importance(best_lgb$best_model)
lgb_importance_sorted <- lgb_importance[order(-lgb_importance$Gain), ]

# keep top n important features
topN <- 10
lgb_top_importance <- lgb_importance_sorted[1:topN, ]

# descriptions of top 10 features
lgb_feature_descriptions <- c(
  "Batting average – hits per at-bat",                          # AVG
  "Strikeout percentage – proportion of plate appearances ending in strikeout",  # K.Percent
  "Intentional walks – plate appearances ending in intentional walk",            # IBB
  "Strikeouts – total number of strikeouts",                     # SO
  "Batting average on balls in play – luck-independent hitting measure",         # BABIP
  "Weighted runs above average – overall offensive contribution",               # wRAA
  "Isolated power – extra base hits per at-bat",                 # ISO
  "Sacrifice flies – fly balls scored for a run",               # SF
  "Grounded into double play – number of double plays hit into", # GDP
  "At-bats – total plate appearances minus walks, hits, etc."   # AB
)

lgb_top_importance$Descriptions <- lgb_feature_descriptions

# Print nicely in RMarkdown
kable(
  lgb_top_importance,
  digits = 4,
  caption = sprintf("LightGBM - Top %d Features by Gain", topN),
  align = c("l", "r", "r", "r")  # left align Feature, right align numbers
)
```

# Comparison of Model Classes

Here we evaluate the best model from each of the previous sections using performance diagnostics and XAI techniques.

## Test Set Evaluation

We evaluate on the test data and calculate RMSE, R-Squared, and plot one-to-one plots.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#Prepare test data same columns as training
drop_cols <- c("Season", "Name", "Team", "next_AVG")
x_test <- as.matrix(test.set[, !(names(test.set) %in% drop_cols)])
y_test <- test.set$next_AVG

#Metrics
rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))
r2   <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)

#One-to-one plot function
one_to_one_plot <- function(actual, pred, title_text) {
  df <- data.frame(Actual = actual, Predicted = pred)
  ggplot(df, aes(x = Actual, y = Predicted)) +
    geom_point(alpha = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    scale_x_continuous(limits=c(0.125,0.375)) +
    scale_y_continuous(limits=c(0.125,0.375)) +
    coord_fixed(ratio=1) +
    labs(
      title = title_text,
      x = "Actual next_AVG",
      y = "Predicted next_AVG"
    ) +
    theme_minimal()
}

# 1) Regularized Regression (glmnet)
pred_lr <- as.numeric(predict(best_model_lr, s = "lambda.min", newx = x_test))


# 2) LightGBM
pred_lgb <- as.numeric(predict(best_lgb$best_model, x_test,
                               num_iteration = best_lgb$best_model$best_iter))

#Results table
results_test <- data.frame(
  Model = c("Regularized Regression (glmnet)", "LightGBM"),
  RMSE = c(rmse(y_test, pred_lr), rmse(y_test, pred_lgb)),
  R_Squared = c(r2(y_test, pred_lr), r2(y_test, pred_lgb))
)

knitr::kable(results_test, digits = 5, caption = "Test Set Performance (RMSE and R-Squared)")

#Plots
p_lr  <- one_to_one_plot(y_test, pred_lr,  "Test Set: glmnet")
p_lgb <- one_to_one_plot(y_test, pred_lgb, "Test Set: LightGBM")

combined_plot <- p_lr + p_lgb + plot_layout(ncol = 3)

#Save plots
ggsave("test_one_to_one_glmnet.png",  p_lr,  width = 6, height = 5)
ggsave("test_one_to_one_lightgbm.png",p_lgb, width = 6, height = 5)

```

```{r one-to-one-pdf, echo=FALSE, fig.width=6, fig.height=5, fig.align='center'}
combined_plot
```

## XAI / IML Analysis

We examine model-agnostic explainability results on the *test set*. This includes global permutation feature importance, partial dependence feature effects, and microscopic SHAP.

```{r xai_iml_test, echo=FALSE, message=FALSE, warning=FALSE}
# ---------------------------
# Prepare TEST-set predictor objects (iml)
# ---------------------------
drop_cols <- c("Season", "Name", "Team", "next_AVG")
x_test <- as.matrix(test.set[, !(names(test.set) %in% drop_cols)])
y_test <- test.set$next_AVG
X_test_df <- as.data.frame(x_test)

# predict functions for iml (glmnet + LightGBM only)
predict_lr_fun <- function(model, newdata) {
  as.numeric(predict(model, s = "lambda.min", newx = as.matrix(newdata)))
}

predict_lgb_fun <- function(model, newdata) {
  # model is a lightgbm Booster; use best iteration
  as.numeric(predict(model, as.matrix(newdata), num_iteration = model$best_iter))
}

# iml predictors on TEST set (same X for both models)
predictor_lr_test <- Predictor$new(
  model = best_model_lr,
  data = X_test_df,
  y = y_test,
  predict.fun = predict_lr_fun
)

predictor_lgb_test <- Predictor$new(
  model = best_lgb$best_model,
  data = X_test_df,
  y = y_test,
  predict.fun = predict_lgb_fun
)

# ---------------------------
# Compute predictions + errors on TEST set
# ---------------------------
pred_test_lr <- predict_lr_fun(best_model_lr, X_test_df)
pred_test_lgb <- predict_lgb_fun(best_lgb$best_model, X_test_df)

abs_err_lr <- abs(y_test - pred_test_lr)
abs_err_lgb <- abs(y_test - pred_test_lgb)

# ---------------------------
# Global feature importance (Permutation Importance)
# ---------------------------
set.seed(123)
imp_lr_test <- FeatureImp$new(predictor_lr_test, loss = "mse", n.repetitions = 10)
imp_lgb_test <- FeatureImp$new(predictor_lgb_test, loss = "mse", n.repetitions = 10)

# Top-10 importance comparison table
top10_lr <- head(imp_lr_test$results[order(-imp_lr_test$results$importance), c("feature","importance")], 10)
top10_lgb <- head(imp_lgb_test$results[order(-imp_lgb_test$results$importance), c("feature","importance")], 10)

top10_lr$importance <- round(top10_lr$importance, 4)
top10_lgb$importance <- round(top10_lgb$importance, 4)

imp_compare <- data.frame(
  Rank = 1:10,
  LR_Feature = top10_lr$feature,
  LR_Importance = top10_lr$importance,
  LGB_Feature = top10_lgb$feature,
  LGB_Importance = top10_lgb$importance
)

# ---------------------------
# Microscopic explanations (Shapley) for extreme predictions on TEST set
# ---------------------------
rmse_fn <- function(y, yhat) sqrt(mean((y - yhat)^2))
best_model_name <- if (rmse_fn(y_test, pred_test_lr) <= rmse_fn(y_test, pred_test_lgb)) "LR" else "LGB"

# pick extremes based on LightGBM predictions + worst-case from best model
idx_high <- which.max(pred_test_lgb)
idx_low  <- which.min(pred_test_lgb)
idx_worst <- if (best_model_name == "LR") which.max(abs_err_lr) else which.max(abs_err_lgb)

case_indices <- unique(c(idx_high, idx_low, idx_worst))
case_indices <- case_indices[1:3]

case_tbl <- data.frame(
  Case = c("Highest predicted (LGBM)", "Lowest predicted (LGBM)", paste0("Largest error (", best_model_name, ")")),
  Index = case_indices,
  Actual = round(y_test[case_indices], 4),
  Pred_LR  = round(pred_test_lr[case_indices], 4),
  Pred_LGB = round(pred_test_lgb[case_indices], 4),
  AbsErr_LR  = round(abs_err_lr[case_indices], 4),
  AbsErr_LGB = round(abs_err_lgb[case_indices], 4)
)
```

### Microscopic

Here we examine the predictions, errors, and actual values for a set of selected observations before examining the SHAP values [@lundberg2017unified] of those instances.

```{r table-of-observations, echo=FALSE}
kable(case_tbl, caption = "Chosen test-set cases for SHAP explanations")
```

```{r micro-shap, echo=FALSE, fig.width=7, fig.height=3}
set.seed(123)
for (idx in case_indices) {
  x_interest <- X_test_df[idx, , drop = FALSE]

  shap_lr <- Shapley$new(predictor_lr_test, x.interest = x_interest, sample.size = 100)
  shap_lgb <- Shapley$new(predictor_lgb_test, x.interest = x_interest, sample.size = 100)

  p_lr  <- shap_lr$plot() + 
           ggtitle(paste("LR - index", idx)) + 
           theme_minimal(base_size = 8) +               
           theme(plot.margin = margin(2, 2, 2, 2))
           
  p_lgb <- shap_lgb$plot() + 
           ggtitle(paste("LGBM - index", idx)) + 
           theme_minimal(base_size = 8) + 
           theme(plot.margin = margin(2, 2, 2, 2))

  invisible(grid.arrange(p_lr, p_lgb, ncol = 2))
}

```

### Macroscopic
Here we plot the Partial Dependence Plots [@hastie2009] for two chosen features.
```{r pdp-plots, echo=FALSE, fig.height=6, fig.width=6, fig.align='left'}
# Select top features for PDP
feat_candidates <- unique(c(top10_lr$feature[1:5], top10_lgb$feature[1:5]))
pdp_features <- feat_candidates[1:2]

pdp_plots <- list()
for (i in seq_along(pdp_features)) {
  f <- pdp_features[i]
  fe_lr  <- FeatureEffect$new(predictor_lr_test,  feature=f, method="pdp", grid.size=20)
  fe_lgb <- FeatureEffect$new(predictor_lgb_test, feature=f, method="pdp", grid.size=20)
  
  pdp_plots[[2*i-1]] <- plot(fe_lr) + ggtitle(paste("PDP - LR:", f)) + theme_minimal(base_size=8)
  pdp_plots[[2*i]]   <- plot(fe_lgb)+ ggtitle(paste("PDP - LGBM:", f)) + theme_minimal(base_size=8)
}

do.call(grid.arrange, c(pdp_plots, ncol=2))
```

Here we plot the permutation feature importances for the two models. Note that the LightGBM importances agree with the model specific ones for the first few features. The regularized regression explanation technique disagrees aside from the K percent. Also note that K percent is ranked high in both models' feature importance plots.
```{r perm-feat-importances, echo = FALSE, fig.width=6, fig.height=5, fig.align='left'}
p_imp_lr  <- plot(imp_lr_test)  + ggtitle("Regularized Regression") + theme_minimal(base_size=8)
p_imp_lgb <- plot(imp_lgb_test) + ggtitle("LightGBM") + theme_minimal(base_size=8)
grid.arrange(p_imp_lr, p_imp_lgb, ncol=2)
```

# References





