---
title: "NickContributions"
author: "Nick Chandler"
date: "2025-12-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(dplyr)
library(corrplot)
library(glmnet)
library(xgboost)
library(iml)
library(dplyr)
library(lightgbm)
library(ggplot2)
```

## Introduction
In this report, we examine the application of two statistical learning methods to baseball data from the years 2020-2024. Specifically we are interested in predicting the next season's batting average of arbitrary players given the standard and advanced metrics of the previous season from this [source](). The report begins with a quick exploratory data analysis (EDA), it then moves into a mathematical description of the statistical learning methods applied, then the application of each method, and finally a performance evaluation and description. This report illustrates how classical, interpretable methods can be applied, and then how more modern (and less interpretable) methods can be applied for better performance, and with the help of XAI, also be interpretable.

## EDA
Here we do a short exploratory analysis of the data to understand some of the basic characteristics.

```{r, echo=FALSE}
data <- read.csv("merged.csv")
```

```{r, echo=FALSE}
num_unique_players <- data %>%
  distinct(Name) %>%       # keep only unique player names
  nrow()                   # count them
```
The dataset has `r nrow(data)` rows, `r ncol(data)` columns, and contains `r num_unique_players` unique players.

```{r, echo=FALSE}

descriptions <- c(
"Season of the game",
"Player name",
"Team abbreviation",
"Games played",
"At bats",
"Hits",
"Singles",
"Doubles",
"Triples",
"Home runs",
"Runs scored",
"RBIs",
"Walks",
"Intentional walks",
"Strikeouts",
"Hit by pitch",
"Sacrifice flies",
"Sacrifice hits",
"Grounded into double plays",
"Stolen bases",
"Caught stealing",
"Plate appearances",
"Walk percentage",
"Strikeout percentage",
"Walk/Strikeout ratio",
"Batting average",
"On-base percentage",
"Slugging percentage",
"OPS",
"Isolated power",
"Speed score",
"BABIP",
"Ultimate base running",
"Weighted GDP",
"Expected runs (XBR)",
"Weighted SB",
"Weighted runs created",
"Weighted RAA",
"Weighted OBA",
"Weighted RC%",
"Next season AVG"
)
head_data <- head(data)
head_data_with_desc <- rbind(descriptions, head_data)
kable(t(head_data_with_desc), caption = "Transposed view of first 6 rows")
```

```{r, echo = FALSE}
players_per_season <- data %>%
  distinct(Name, Season, .keep_all = TRUE) %>%  # drop duplicates if needed
  group_by(Season) %>%
  summarise(Number_of_Players = n_distinct(Name)) %>%
  arrange(Season)

# Render as a clean table
kable(t(players_per_season), caption = "Total number of players in each season")
```



Correlation plots using Pearson's correlation coefficient.
```{r, echo = FALSE}
numeric_data <- data[, sapply(data, is.numeric)]

cor_matrix_pearson <- cor(numeric_data, use = "pairwise.complete.obs", method = "pearson")
corrplot(cor_matrix_pearson, method = "color", tl.cex = 0.6)
```

## Mathematical Description of Regularized Regression

NOTE TO SELF: All equations can be derived from the ESL textbook

This section examines the mathematical background for the regularized regression model types of: Ridge, Lasso, and Elastic-Net regression. All assume a model of the form (here, without interaction terms):
$$y = \beta_0 + \sum_{i=1}^{p} \beta_i x_i + \varepsilon.$$
Where the $\beta_i$s are found by solving (though specific instances introduce additional terms):
$$\min_\vec{\beta} \left\|  X\vec{\beta} - \vec{y} \right\|^2_2\text{, with } X \in \mathbb{R}^{n \times p} \text{ the design matrix}$$
The general solution is given (analytically) by:
$$\hat{\vec{\beta}} = \left(X^T X \right)^{-1} X^T\vec{y}$$
We now explain the various cases of regularization.

### Ridge Regression
The Ridge regression penalizes the size of the parameters with an L2 norm. Specifically the optimization problem is given by,
$$\min_{\vec{\beta}} \left\| X\vec{\beta} - \vec{y} \right\|_2^2 + \lambda \| \vec{\beta} \|_2^2$$

The analytical solution is given by:
$$ \vec{\beta} = \left(  X^TX + \lambda I \right)^{-1}X^T\vec{y} $$
Note that in the previous two equations, $\lambda$ is a hyperparameter that can be tuned with cross validation. We now proceed to Lasso Regression.

### Lasso Regression
The Lasso regression penalizes the size of the parameters with an L1 norm. The optimization problem is:
$$\min_{\vec{\beta}} \left\| X\vec{\beta} - \vec{y} \right\|_2^2 + \lambda \| \vec{\beta} \|_1$$
Since the L1 norm is not differentiable at $0$, there is no closed-form, analytical solution to the above problem. One can use solvers for convex problems capable of dealing with non-smooth objectives. The optimality condition is given by: 

$$
0 \in 2 X^\top (X \hat{\beta} - y) + \lambda\, \partial \|\hat{\beta}\|_1,
\quad
(\partial \|\hat{\beta}\|_1)_i =
\begin{cases}
\{1\}, & \hat{\beta}_i > 0,\\[1mm]
\{-1\}, & \hat{\beta}_i < 0,\\[1mm]
[-1,1], & \hat{\beta}_i = 0
\end{cases}
$$

Again, $\lambda$ is a hyperparameter that can be tuned with cross validation. We now move onto the generalization of the previous two regularization methods, elastic-net regression.

### Elastic-Net Regression
The elastic-net regression uses a convex combination of the lasso penalty (L1) and ridge penalty (L2). Specifically, the optimization problem is given by:
$$\min_{\vec{\beta}} \left\| X\vec{\beta} - \vec{y} \right\|_2^2 + \lambda \left( \left(1 - \alpha  \right) \| \vec{\beta} \|_1 + \alpha\|\vec{\beta}  \|_2^2 \right)$$

We again have the L1 norm which is not differentiable at $0$. Thereby a convex solver capable of optimizing non-smooth objectives should be used. The optimality condition is given by:
Since the L1 norm is not differentiable at $0$, there is no closed-form, analytical solution to the above problem. Once again solvers for convex problems capable of dealing with non-smooth objectives must be applied. The optimality condition is given by: 

$$
0 \in 2 X^\top (X \hat{\beta} - y) +  \lambda \left( 2 \alpha \hat{\vec{\beta}} + (1 - \alpha) \partial \|\hat{\beta}\|_1 \right),
\quad
(\partial \|\hat{\beta}\|_1)_i =
\begin{cases}
\{1\}, & \hat{\beta}_i > 0,\\[1mm]
\{-1\}, & \hat{\beta}_i < 0,\\[1mm]
[-1,1], & \hat{\beta}_i = 0
\end{cases}
$$

The parameters $\lambda$ and $\alpha$ are hyperparameters to be tuned via cross validation.


## Mathematical Description of Gradient Boosting Decesion Trees
In this section we give a short mathematical description of Gradient Boosting Decision Trees (focusing on the gradient boosting aspect). This does not include complexities introduced with runtime optimization as in XGBoost.

Suppose our training data is given by: $\left\{ (\vec{x_i}, y_i) \right\}_{i=1}^n$

Gradient boosting builds an additive sequential model interatively, aiming to minimize a loss function: $L(y, F(x))$. Specifically, gradient boosting builds a model (defined recursively):
$$F_m(x) = F_{m-1}(x) + \eta \cdot \gamma_m \cdot h_m(x)$$
with,
$$F_0(x) = \underset{\gamma}{\arg \min} \sum_{i=1}^{n} L (y_i, \gamma)$$
The two step process (for each iteration) to fit the model consists of:
1. Building weak learners $h_m(x)$.
2. Finding the optimal step size at that iteration $\gamma_m$.

For step 1. we must compute the pseudo residual: 
$$\tilde{y}_{im} = - \left[ \frac{ \partial L(y_i, F(x_i)) }{ \partial F(x_i) } \right]_{F(x) = F_{m-1}(x)}$$
and then fit the wewak learner $h_m(x)$ (in our case, a tree model) to the given pseudo residual. 

We then do step 2. to find the optimal step size $\gamma_m$ using an optimizer to solve:
$$\underset{\gamma}{\arg \min} \sum_{i = 1}^{n} L(y_i, F_{m-1}(x_i) + \gamma h_m(x_i))$$
Finally, the $\eta$ given in the original equation is a hyperparameter and must be tuned.





## Train-Val-Test Split

We split the data into a train set (60%), validation (val) set (20%), and a test set (20%). Specifically, we adhere to the following condition when deciding on the splits: A player may occur in no more than one set. Additionally, the newest data (that from the 2024 season) is saved for the test set. In numbers:
```{r, echo=FALSE}
# All unique players
players <- unique(data$Name)

## PLAN: Take all 2024-appearing players and then more to get the 20% necessary for test. Then just split the val and train normally but with players not occurring in two sets.

# Players in 2024 season -> all go to test
players.2024 <- unique(data[data$Season == 2024,"Name"])
players.remaining <- setdiff(players, players.2024)  # setdiff -> set difference

# How many players to get 20% of rows
rows.2024 <- data[data$Name %in% players.2024, ]
target.test.rows <- floor(0.2 * nrow(data))
extra.needed <- target.test.rows - nrow(rows.2024)

# sample extra players from remaining players
extra.players <- c()
cum.rows <- 0
set.seed(123)
shuffled <- sample(players.remaining)  # Shuffles the players.remaining list

for (p in shuffled) {
  player.rows <- nrow(data[data$Name == p, ])  # Take out the player's rows
  cum.rows <- cum.rows + player.rows  # Add to the count of the cumulative rows
  extra.players <- c(extra.players, p)  # Add new player to the extra players
  if (cum.rows >= extra.needed) break  # Stop if we got the needed players (won't overshoot 20% by much)
}

test.players <- c(players.2024, extra.players)
test.set <- na.omit(data[data$Name %in% test.players, ])

trainval.players <- setdiff(players, test.players)
trainval.data <- data[data$Name %in% trainval.players, ]

# val - 20%
val.frac <- 0.247  # MAGIC NUMBER (kind of, at least there's a var name) !
val.players <- sample(unique(trainval.data$Name),
                      size = ceiling(val.frac * length(unique(trainval.data$Name))))
val.set <- na.omit(trainval.data[trainval.data$Name %in% val.players, ])

# Train is the rest
train.set <- na.omit(trainval.data[!trainval.data$Name %in% val.players, ])


```
- The train set has `r nrow(train.set)` rows and `r length(unique(train.set$Name))` unique players.  
- The validation set has `r nrow(val.set)` rows and `r length(unique(val.set$Name))` unique players.
- The test set has `r nrow(test.set)` rows and `r length(unique(test.set$Name))` unique players.  


## Regularized Regression Application

In this section, we apply the regularized regression techniques.


```{r}
drop_cols <- c("Season", "Name", "Team", "next_AVG")

x_train <- as.matrix(train.set[, !(names(train.set) %in% drop_cols)])
y_train <- train.set$next_AVG

x_val <- as.matrix(val.set[, !(names(val.set) %in% drop_cols)])
y_val <- val.set$next_AVG

# Define the alpha sets
ridge_alpha <- 0
lasso_alpha <- 1
elastic_alphas <- seq(0.1, 0.9, by = 0.1)  # elastic net grid (exclude 0 and 1)

# Store results
results <- list()

# Function to train and evaluate
train_eval <- function(alpha_value) {
  cv_fit <- cv.glmnet(x_train, y_train, alpha = alpha_value, nfolds = 5)  # Be able to explain cross validation
  pred_val <- predict(cv_fit, s = "lambda.min", newx = x_val)  # You must understand this well - WTH is lambda.min, what exactly does it do?
  mse_val <- mean((y_val - pred_val)^2)
  list(model = cv_fit, mse = mse_val, alpha = alpha_value)
}

# 1) Ridge
results[["ridge"]] <- train_eval(ridge_alpha)

# 2) Lasso
results[["lasso"]] <- train_eval(lasso_alpha)

# 3) Elastic Net Grid
for (a in elastic_alphas) {
  key <- paste0("elastic_alpha_", a)
  results[[key]] <- train_eval(a)
}

# Find best model overall
best_key_lr <- names(results)[which.min(sapply(results, function(x) x$mse))]
best_model_lr <- results[[best_key_lr]]$model
best_mse_lr <- round(results[[best_key_lr]]$mse, 3)
```

Through this analysis, we find that `r best_key_lr` is attains the minimum validation RMSE of all models tested at (rounded): `r round(sqrt(best_mse_lr),6)`. Which means that the model mispredicts the next season's batting average by approximately `r round(sqrt(best_mse_lr)*100,4)`% on average. We examine the best model further, using classical diagnostics and model-agnostic XAI techniques.

The following is a plot of the lambda value in `r best_key_lr` regression.
```{r, echo=FALSE}
plot(best_model_lr)
```

The top 10 coefficients (ranked in order of magnitude) are as follows:
```{r, echo=FALSE}
# Get coefficients at best lambda
coef_best <- coef(best_model_lr, s = "lambda.min")

# Convert to data frame
coef_df <- data.frame(
  Feature = rownames(coef_best),
  Coefficient = as.numeric(coef_best)
)

# Remove intercept
coef_df <- coef_df[coef_df$Feature != "(Intercept)", ]

# Sort by absolute magnitude
coef_df$Magnitude <- abs(coef_df$Coefficient)
coef_df <- coef_df[order(-coef_df$Magnitude), c("Feature", "Coefficient")]

# Optional: round for reporting
coef_df$Coefficient <- round(coef_df$Coefficient, 3)

# Keep top 10 contributors
top_coef_df <- head(coef_df, 10)

# Add descriptions
top_coef_df$Description <- c(
  "Batting average – hits per at-bat",
  "Strikeout percentage – proportion of plate appearances ending in strikeout",
  "Walk percentage – proportion of plate appearances ending in walk",
  "Isolated power – extra base hits per at-bat",
  "Batting average on balls in play – luck-independent hitting measure",
  "On-base percentage – times reaching base per plate appearance",
  "On-base plus slugging – overall offensive contribution",
  "Weighted on-base average – combines OBP and power into single metric",
  "Slugging percentage – total bases per at-bat",
  "Walk-to-strikeout ratio – plate discipline indicator"
)

# Render table without row names
kable(top_coef_df, caption = "Top 10 Ridge Regression Coefficients with Descriptions", row.names = FALSE)
```

So as we see the most important coefficients are the largest when using regularized regression. It can be reasoned that they contribute more to the reduction of data fit error (Sum of Squared Errors) than they increase the total loss with the penalization of size, indicating that they are important in predicting the response. We examine the effect of the parameters on the predictions again using SHAP in the final comparison section.



## Gradient Boosted Decision Trees Application

In this section, we apply the gradient boosted decision trees. Specifically we examine the implementations: CatBoost, LightGBM, and XGBoost.

VERY IMPORTANT
README BEFORE BEGINNING:

REMOVE THE "Season", "Name", AND, "Team" COLUMNS BEFORE TRAINING!!!!

DO NOT USE THE TEST SET DURING YOUR ANALYSES - THIS WILL BE USED AT THE END

```{r, warning=FALSE}
# Drop columns and prepare matrices
drop_cols <- c("Season", "Name", "Team", "next_AVG")

x_train_xgb <- as.matrix(train.set[, !(names(train.set) %in% drop_cols)])
y_train_xgb <- train.set$next_AVG

x_val_xgb <- as.matrix(val.set[, !(names(val.set) %in% drop_cols)])
y_val_xgb <- val.set$next_AVG

# Define hyperparameter grids
learning_rates <- c(0.001, 0.01, 0.05, 0.1, 0.2)
subsamples <- c(0.5, 0.7, 1.0)
ntrees <- c(50, 100, 200, 500)

# Store results
xgb_results <- list()

# Function to train and evaluate
train_eval_xgb <- function(eta_val, subsample_val, nrounds_val) {
  dtrain <- xgb.DMatrix(data = x_train_xgb, label = y_train_xgb)
  dval <- xgb.DMatrix(data = x_val_xgb, label = y_val_xgb)
  
  params <- list(
    objective = "reg:squarederror",  # regression
    eta = eta_val,
    subsample = subsample_val,
    max_depth = 6,                   # fixed for simplicity, can tune later
    eval_metric = "rmse"
  )
  
  xgb_fit <- xgb.train(
    params = params,
    data = dtrain,
    nrounds = nrounds_val,
    watchlist = list(val = dval),
    verbose = 0
  )
  
  pred_val <- predict(xgb_fit, dval)
  mse_val <- mean((y_val_xgb - pred_val)^2)
  
  list(model = xgb_fit, mse = mse_val, eta = eta_val,
       subsample = subsample_val, nrounds = nrounds_val)
}

# Grid search
for (eta_val in learning_rates) {
  for (sub_val in subsamples) {
    for (n_val in ntrees) {
      key <- paste0("xgb_eta_", eta_val, "_sub_", sub_val, "_ntree_", n_val)
      xgb_results[[key]] <- train_eval_xgb(eta_val, sub_val, n_val)
    }
  }
}

# Find best model overall
best_key_xgb <- names(xgb_results)[which.min(sapply(xgb_results, function(x) x$mse))]
best_model_xgb <- xgb_results[[best_key_xgb]]$model
best_mse_xgb <- round(xgb_results[[best_key_xgb]]$mse, 3)

```

Through this analysis, we find that `r best_key_xgb` attains the minimum validation RMSE of all XGBoost models tested at (rounded): `r round(sqrt(best_mse_xgb),6)`. This indicates that, on average, the model mispredicts the next season's target by approximately `r round(sqrt(best_mse_xgb)*100,4)`%. This is similar to the Regularized Regression model, perhaps indicating that linear relationships between the features and target variable are most prevalent. 
We proceed to examine the best XGBoost model further using feature importance


### Feature Importance (XGB-Specific)

```{r, echo=FALSE}
feature_names <- colnames(x_train_xgb)

importance_matrix <- xgb.importance(feature_names=feature_names, model=best_model_xgb)

top_features <- head(importance_matrix, 10)
  top_features_df <- data.frame(
  Feature = top_features$Feature,
  Gain = round(top_features$Gain, 3),
  Cover = round(top_features$Cover, 3),
  Frequency = round(top_features$Frequency, 3)
)

top_features_df$Description <- c(
  "Batting average – hits per at-bat",
  "Strikeout percentage – proportion of plate appearances ending in strikeout",
  "Walk percentage – proportion of plate appearances ending in walk",
  "Isolated power – extra base hits per at-bat",
  "Batting average on balls in play – luck-independent hitting measure",
  "On-base percentage – times reaching base per plate appearance",
  "On-base plus slugging – overall offensive contribution",
  "Weighted on-base average – combines OBP and power into single metric",
  "Slugging percentage – total bases per at-bat",
  "Walk-to-strikeout ratio – plate discipline indicator"
)

kable(top_features_df, caption = "Top 10 XGBoost Feature Importances with Descriptions", row.names = FALSE)
```

The columns **`Gain`**, **`Cover`**, and **`Frequency`** in the XGBoost feature importance table have the following meanings (from R help):

- **`Gain`**: Fractional contribution of each feature to the model based on the total gain of this feature’s splits. Higher values indicate more important predictive features.  
- **`Cover`**: Relative number of observations affected by splits involving this feature.  
- **`Frequency`**: Relative number of times a feature is used in the trees. Higher values mean the feature appears more often across all trees.

```{r, echo = FALSE}
max_gain <- importance_matrix[which.max(importance_matrix$Gain), ]
max_cover <- importance_matrix[which.max(importance_matrix$Cover), ]
max_freq <- importance_matrix[which.max(importance_matrix$Frequency), ]

max_features_df <- data.frame(
  Metric = c("Gain", "Cover", "Frequency"),
  Feature = c(max_gain$Feature, max_cover$Feature, max_freq$Feature),
  Score = c(round(max_gain$Gain, 3),
  round(max_cover$Cover, 3),
  round(max_freq$Frequency, 3))
)

max_features_df$Description <- c(
  "Batting average – hits per at-bat",
  "Strikeout percentage – proportion of plate appearances ending in strikeout",
  "Games played – total games in the season"
)

kable(max_features_df, caption = "Features with Maximal Gain, Cover, and Frequency", row.names = FALSE)
```

We again see the last year's average at the top in terms of gain, indicating that it has the best predictive ability for predicting the next year's average when considering our models.

### LightGBM
```{r, echo=FALSE}
# lightgbm training and evaluation function
train_eval_lgb <- function(x_train, y_train, x_val, y_val, params, nrounds=5000, early_stop=100) {
  
  dtrain <- lgb.Dataset(data = x_train, label = y_train)
  dval   <- lgb.Dataset(data = x_val, label = y_val)
  
  evals <- list(train = dtrain, val = dval)
  
  model <- lgb.train(
    params = params,
    data = dtrain,
    nrounds = nrounds,
    valids = evals,
    early_stopping_rounds = early_stop,
    verbose = -1
  )
  
  # Predictions on validation set
  preds <- predict(model, x_val, num_iteration = model$best_iter)
  rmse_val <- sqrt(mean((preds - y_val)^2))
  
  # Record train/val RMSE for plotting
  train_rmse <- sapply(model$record_evals$train$rmse$eval, identity)
  val_rmse   <- sapply(model$record_evals$val$rmse$eval, identity)
  
  return(list(
    model = model,
    rmse_val = rmse_val,
    best_iter = model$best_iter,
    train_rmse = train_rmse,
    val_rmse = val_rmse
  ))
}

# function for plotting train/val curves
plot_rmse_curves <- function(train_rmse, val_rmse) {
  df_plot <- data.frame(
    Iteration = 1:length(train_rmse),
    Train_RMSE = train_rmse,
    Val_RMSE = val_rmse
  )
  
  ggplot(df_plot, aes(x = Iteration)) +
    geom_line(aes(y = Train_RMSE, color = "Train")) +
    geom_line(aes(y = Val_RMSE, color = "Validation")) +
    labs(y = "RMSE", title = "Train vs Validation RMSE") +
    scale_color_manual(values = c("Train"="blue", "Validation"="red")) +
    theme_minimal()
  
}

#function for hyperparamter optimization
hpo_lgb <- function(x_train, y_train, x_val, y_val, param_grid, nrounds=5000, early_stop=100) {
  
  best_rmse <- Inf
  best_params <- NULL
  best_model <- NULL
  best_curves <- NULL
  
  for (i in 1:nrow(param_grid)) {
    
    params <- as.list(param_grid[i, ])
    params$objective <- "regression"
    params$metric <- "rmse"
    
    res <- train_eval_lgb(x_train, y_train, x_val, y_val, params, nrounds, early_stop)
    
    # Print both train and validation RMSE
  #cat(sprintf(
  #  "Grid %d/%d --> Train RMSE: %.4f | Val RMSE: %.4f\n",
  #  i, nrow(param_grid),
  #  tail(res$train_rmse, 1),  # last train RMSE (best iteration)
  #  res$rmse_val
  #))
    
    if (res$rmse_val < best_rmse) {
      best_rmse <- res$rmse_val
      best_params <- params
      best_model <- res$model
      best_curves <- list(train = res$train_rmse, val = res$val_rmse)
    }
  }
  
  return(list(
    best_model = best_model,
    best_params = best_params,
    best_rmse = best_rmse,
    train_rmse = best_curves$train,
    val_rmse = best_curves$val
  ))
}

print_best_lgb <- function(hpo_result) {
  
  cat("=====================================\n")
  cat("          Best LightGBM Model        \n")
  cat("=====================================\n")
  
  cat(sprintf("Best Validation RMSE : %.4f\n", hpo_result$best_rmse))
  cat(sprintf("Best Training RMSE   : %.4f\n", tail(hpo_result$train_rmse, 1)))
  cat(sprintf("Best Iteration       : %d\n", hpo_result$best_model$best_iter))
  cat("\n")
  cat("Selected Hyperparameters:\n")
  
  # Print each parameter neatly
  for (param_name in names(hpo_result$best_params)) {
    cat(sprintf("  %-20s : %s\n", param_name, hpo_result$best_params[[param_name]]))
  }
  
  cat("=====================================\n")
}
```

```{r, echo=FALSE}
# parameter search grid for HPO
param_grid <- expand.grid(
  learning_rate = c(0.001, 0.005, 0.01),
  max_depth = c(5,6,7),
  num_leaves = c(8,16),
  feature_fraction = c(0.8, 1)
)


# run the hpo
best_lgb <- hpo_lgb(x_train, y_train, x_val, y_val, param_grid, nrounds = 1000, early_stop = 50)

# plot train/val curves
plot_rmse_curves(best_lgb$train_rmse, best_lgb$val_rmse)

# print best results
print_best_lgb(best_lgb)
```
### LightGBM Feature Importance

```{r, echo=FALSE}
# feature importance
lgb_importance <- lgb.importance(best_lgb$best_model)
lgb_importance_sorted <- lgb_importance[order(-lgb_importance$Gain), ]

# keep top n important features
topN <- 10
lgb_top_importance <- lgb_importance_sorted[1:topN, ]

# descriptions of top 10 features
lgb_feature_descriptions <- c(
  "Batting average – hits per at-bat",                          # AVG
  "Strikeout percentage – proportion of plate appearances ending in strikeout",  # K.Percent
  "Intentional walks – plate appearances ending in intentional walk",            # IBB
  "Strikeouts – total number of strikeouts",                     # SO
  "Batting average on balls in play – luck-independent hitting measure",         # BABIP
  "Weighted runs above average – overall offensive contribution",               # wRAA
  "Isolated power – extra base hits per at-bat",                 # ISO
  "Sacrifice flies – fly balls scored for a run",               # SF
  "Grounded into double play – number of double plays hit into", # GDP
  "At-bats – total plate appearances minus walks, hits, etc."   # AB
)

lgb_top_importance$Descriptions <- lgb_feature_descriptions

# Print nicely in RMarkdown
kable(
  lgb_top_importance,
  digits = 4,
  caption = sprintf("LightGBM - Top %d Features by Gain", topN),
  align = c("l", "r", "r", "r")  # left align Feature, right align numbers
)

lgb.plot.importance(lgb_importance)
```

## Comparison of Model Classes

Here we evaluate the best model from each of the previous sections.

### Test Set Evaluation

We evaluate on the test data



### SHAP Analysis

We examine the SHAP values for each of the best models and specifically focus on a few microscopic explanations of the extreme predictions.


```{r, message=FALSE, warning=FALSE}


# ---------------------------
# Prepare predictor objects
# ---------------------------

# For glmnet / regularized regression
# iml expects a data.frame for X and a predict function that returns numeric vector
X_val_lr <- as.data.frame(x_val)  

predict_lr <- function(model, newdata) {
  # Returns a numeric vector of predictions
  as.numeric(predict(model, s = "lambda.min", newx = as.matrix(newdata)))
}

predictor_lr <- Predictor$new(
  model = best_model_lr,
  data = X_val_lr,
  y = y_val,
  predict.fun = predict_lr
)

# For XGBoost
X_val_xgb_df <- as.data.frame(x_val_xgb)

predict_xgb <- function(model, newdata) {
  predict(model, as.matrix(newdata))
}

predictor_xgb <- Predictor$new(
  model = best_model_xgb,
  data = X_val_xgb_df,
  y = y_val_xgb,
  predict.fun = predict_xgb
)

# ---------------------------
# Compute global SHAP / feature importance
# ---------------------------

# Use FeatureImp (permutation-based importance) or Shapley for exact contributions
# For global diagnostics, permutation importance is faster
set.seed(123)

imp_lr <- FeatureImp$new(predictor_lr, loss = "mse")
imp_xgb <- FeatureImp$new(predictor_xgb, loss = "mse")

# ---------------------------
# Plot global importance
# ---------------------------
plot(imp_lr) + ggtitle("Regularized Regression - Global Feature Importance")
plot(imp_xgb) + ggtitle("XGBoost - Global Feature Importance")

# ---------------------------
# Optional: exact Shapley values (slower)
# ---------------------------
# Example for first 10 features or subset of validation set
shap_lr <- Shapley$new(predictor_lr, x.interest = X_val_lr[1, ])
shap_lr$plot()  # plots feature contributions for this row

shap_xgb <- Shapley$new(predictor_xgb, x.interest = X_val_xgb_df[1, ])
shap_xgb$plot()


```



## Deployment of Best Model

We examine a few 2025 season statistics to get predictions for our favorite players in the 2026 season. The results won't be available until next year!






