---
title: "NickContributions"
author: "Nick Chandler"
date: "2025-12-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(dplyr)
library(corrplot)
```

## EDA
Here we do a short exploratory analysis of the data to understand some of the basic characteristics.

```{r, echo=FALSE}
data <- read.csv("merged.csv")
```

```{r, echo=FALSE}
num_unique_players <- data %>%
  distinct(Name) %>%       # keep only unique player names
  nrow()                   # count them
```
The dataset has `r nrow(data)` rows, `r ncol(data)` columns, and contains `r num_unique_players` unique players.

```{r, echo=FALSE}

descriptions <- c(
"Season of the game",
"Player name",
"Team abbreviation",
"Games played",
"At bats",
"Hits",
"Singles",
"Doubles",
"Triples",
"Home runs",
"Runs scored",
"RBIs",
"Walks",
"Intentional walks",
"Strikeouts",
"Hit by pitch",
"Sacrifice flies",
"Sacrifice hits",
"Grounded into double plays",
"Stolen bases",
"Caught stealing",
"Plate appearances",
"Walk percentage",
"Strikeout percentage",
"Walk/Strikeout ratio",
"Batting average",
"On-base percentage",
"Slugging percentage",
"OPS",
"Isolated power",
"Speed score",
"BABIP",
"Ultimate base running",
"Weighted GDP",
"Expected runs (XBR)",
"Weighted SB",
"Weighted runs created",
"Weighted RAA",
"Weighted OBA",
"Weighted RC%",
"Next season AVG"
)
head_data <- head(data)
head_data_with_desc <- rbind(descriptions, head_data)
kable(t(head_data_with_desc), caption = "Transposed view of first 6 rows")
```

```{r, echo = FALSE}
players_per_season <- data %>%
  distinct(Name, Season, .keep_all = TRUE) %>%  # drop duplicates if needed
  group_by(Season) %>%
  summarise(Number_of_Players = n_distinct(Name)) %>%
  arrange(Season)

# Render as a clean table
kable(t(players_per_season), caption = "Total number of players in each season")
```



Correlation plots using Pearson's correlation coefficient.
```{r, echo = FALSE}
numeric_data <- data[, sapply(data, is.numeric)]

cor_matrix_pearson <- cor(numeric_data, use = "pairwise.complete.obs", method = "pearson")
corrplot(cor_matrix_pearson, method = "color", tl.cex = 0.6)
```


## Mathematical Description of Gradient Boosting Decesion Trees
In this section we give a short mathematical description of Gradient Boosting Decision Trees (focusing on the gradient boosting aspect). This does not include complexities introduced with runtime optimization as in XGBoost.

Suppose our training data is given by: $\left\{ (\vec{x_i}, y_i) \right\}_{i=1}^n$

Gradient boosting builds an additive sequential model interatively, aiming to minimize a loss function: $L(y, F(x))$. Specifically, gradient boosting builds a model (defined recursively):
$$F_m(x) = F_{m-1}(x) + \eta \cdot \gamma_m \cdot h_m(x)$$
with,
$$F_0(x) = \underset{\gamma}{\arg \min} \sum_{i=1}^{n} L (y_i, \gamma)$$
The two step process (for each iteration) to fit the model consists of:
1. Building weak learners $h_m(x)$.
2. Finding the optimal step size at that iteration $\gamma_m$.

For step 1. we must compute the pseudo residual: 
$$\tilde{y}_{im} = - \left[ \frac{ \partial L(y_i, F(x_i)) }{ \partial F(x_i) } \right]_{F(x) = F_{m-1}(x)}$$
and then fit the wewak learner $h_m(x)$ (in our case, a tree model) to the given pseudo residual. 

We then do step 2. to find the optimal step size $\gamma_m$ using an optimizer to solve:
$$\underset{\gamma}{\arg \min} \sum_{i = 1}^{n} L(y_i, F_{m-1}(x_i) + \gamma h_m(x_i))$$
Finally, the $\eta$ given in the original equation is a hyperparameter and must be tuned.



## Mathematical Description of Regularized Regression

NOTE TO SELF: All equations can be derived from the ESL textbook

This section examines the mathematical background for the regularized regression model types of: Ridge, Lasso, and Elastic-Net regression. All assume a model of the form (here, without interaction terms):
$$y = \beta_0 + \sum_{i=1}^{p} \beta_i x_i + \varepsilon.$$
Where the $\beta_i$s are found by solving (though specific instances introduce additional terms):
$$\min_\vec{\beta} \left\|  X\vec{\beta} - \vec{y} \right\|^2_2\text{, with } X \in \mathbb{R}^{n \times p} \text{ the design matrix}$$
The general solution is given (analytically) by:
$$\hat{\vec{\beta}} = \left(X^T X \right)^{-1} X^T\vec{y}$$
We now explain the various cases of regularization.

### Ridge Regression
The Ridge regression penalizes the size of the parameters with an L2 norm. Specifically the optimization problem is given by,
$$\min_{\vec{\beta}} \left\| X\vec{\beta} - \vec{y} \right\|_2^2 + \lambda \| \vec{\beta} \|_2^2$$

The analytical solution is given by:
$$ \vec{\beta} = \left(  X^TX + \lambda I \right)^{-1}X^T\vec{y} $$
Note that in the previous two equations, $\lambda$ is a hyperparameter that can be tuned with cross validation. We now proceed to Lasso Regression.

### Lasso Regression
The Lasso regression penalizes the size of the parameters with an L1 norm. The optimization problem is:
$$\min_{\vec{\beta}} \left\| X\vec{\beta} - \vec{y} \right\|_2^2 + \lambda \| \vec{\beta} \|_1$$
Since the L1 norm is not differentiable at $0$, there is no closed-form, analytical solution to the above problem. One can use solvers for convex problems capable of dealing with non-smooth objectives. The optimality condition is given by: 

$$
0 \in 2 X^\top (X \hat{\beta} - y) + \lambda\, \partial \|\hat{\beta}\|_1,
\quad
(\partial \|\hat{\beta}\|_1)_i =
\begin{cases}
\{1\}, & \hat{\beta}_i > 0,\\[1mm]
\{-1\}, & \hat{\beta}_i < 0,\\[1mm]
[-1,1], & \hat{\beta}_i = 0
\end{cases}
$$

Again, $\lambda$ is a hyperparameter that can be tuned with cross validation. We now move onto the generalization of the previous two regularization methods, elastic-net regression.

### Elastic-Net Regression
The elastic-net regression uses a convex combination of the lasso penalty (L1) and ridge penalty (L2). Specifically, the optimization problem is given by:
$$\min_{\vec{\beta}} \left\| X\vec{\beta} - \vec{y} \right\|_2^2 + \lambda \left( \left(1 - \alpha  \right) \| \vec{\beta} \|_1 + \alpha\|\vec{\beta}  \|_2^2 \right)$$

We again have the L1 norm which is not differentiable at $0$. Thereby a convex solver capable of optimizing non-smooth objectives should be used. The optimality condition is given by:
Since the L1 norm is not differentiable at $0$, there is no closed-form, analytical solution to the above problem. Once again solvers for convex problems capable of dealing with non-smooth objectives must be applied. The optimality condition is given by: 

$$
0 \in 2 X^\top (X \hat{\beta} - y) +  \lambda \left( 2 \alpha \hat{\vec{\beta}} + (1 - \alpha) \partial \|\hat{\beta}\|_1 \right),
\quad
(\partial \|\hat{\beta}\|_1)_i =
\begin{cases}
\{1\}, & \hat{\beta}_i > 0,\\[1mm]
\{-1\}, & \hat{\beta}_i < 0,\\[1mm]
[-1,1], & \hat{\beta}_i = 0
\end{cases}
$$

The parameters $\lambda$ and $\alpha$ are hyperparameters to be tuned via cross validation.


## Train-Val-Test Split

We split the data into a train set (60%), validation (val) set (20%), and a test set (20%). Specifically, we adhere to the following condition when deciding on the splits: A player may occur in no more than one set. Additionally, the newest data (that from the 2024 season) is saved for the test set. In numbers:
```{r, echo=FALSE}
# All unique players
players <- unique(data$Name)

## PLAN: Take all 2024-appearing players and then more to get the 20% necessary for test. Then just split the val and train normally but with players not occurring in two sets.

# Players in 2024 season -> all go to test
players.2024 <- unique(data[data$Season == 2024,"Name"])
players.remaining <- setdiff(players, players.2024)  # setdiff -> set difference

# How many players to get 20% of rows
rows.2024 <- data[data$Name %in% players.2024, ]
target.test.rows <- floor(0.2 * nrow(data))
extra.needed <- target.test.rows - nrow(rows.2024)

# sample extra players from remaining players
extra.players <- c()
cum.rows <- 0
set.seed(123)
shuffled <- sample(players.remaining)  # Shuffles the players.remaining list

for (p in shuffled) {
  player.rows <- nrow(data[data$Name == p, ])  # Take out the player's rows
  cum.rows <- cum.rows + player.rows  # Add to the count of the cumulative rows
  extra.players <- c(extra.players, p)  # Add new player to the extra players
  if (cum.rows >= extra.needed) break  # Stop if we got the needed players (won't overshoot 20% by much)
}

test.players <- c(players.2024, extra.players)
test.set <- data[data$Name %in% test.players, ]

trainval.players <- setdiff(players, test.players)
trainval.data <- data[data$Name %in% trainval.players, ]

# val - 20%
val.frac <- 0.247  # MAGIC NUMBER (kind of, at least there's a var name) !
val.players <- sample(unique(trainval.data$Name),
                      size = ceiling(val.frac * length(unique(trainval.data$Name))))
val.set <- trainval.data[trainval.data$Name %in% val.players, ]

# Train is the rest
train.set <- trainval.data[!trainval.data$Name %in% val.players, ]
```
- The train set has `r nrow(train.set)` rows and `r length(unique(train.set$Name))` unique players.  
- The validation set has `r nrow(val.set)` rows and `r length(unique(val.set$Name))` unique players.
- The test set has `r nrow(test.set)` rows and `r length(unique(test.set$Name))` unique players.  


## Gradient Boosted Decision Trees Application

In this section, we apply the gradient boosted decision trees. Specifically we examine the implementations: CatBoost, LightGBM, and XGBoost.


## Regularized Regression Application

In this section, we apply the regularized regression techniques.







